# Daily Bitcoin Time Series Analysis (2014–2025)

by Hans Darmawan

------------------------------------------------------------------------

## 1. Setup & Libraries

```{r}

```

------------------------------------------------------------------------

## 2. Load Daily Bitcoin Data

```{r}

```

### Validation

```{r}

```

------------------------------------------------------------------------

## 3. Data Cleaning & Tsibble Conversion

```{r}

```

### Sanity Checks

```{r}

```

------------------------------------------------------------------------

## 4. Exploratory Data Analysis (EDA)

### 4.1 Full History Price

```{r}

```

------------------------------------------------------------------------

### 4.2 Last 2 Years

```{r}

```

------------------------------------------------------------------------

### 4.3 30-Day Rolling Mean

```{r}

```

------------------------------------------------------------------------

### 4.4 30-Day Rolling Volatility

```{r}

```

------------------------------------------------------------------------

## 5. Baseline Forecasting Models

### 5.1 Train–Test Split

```{r}

```

------------------------------------------------------------------------

### 5.2 Model Fitting

```{r}

```

------------------------------------------------------------------------

### 5.3 Forecasting

```{r}

```

------------------------------------------------------------------------

## 6. Model Evaluation

```{r}

```

### Summary Comparison

```{r}

```

------------------------------------------------------------------------

## 7. Conclusion

-   Dataset successfully modeled as **daily tsibble**

-   Baseline models established:

    -   Naive
    -   Random Walk with Drift
    -   ARIMA

-   Evaluation uses **out-of-sample 14-day horizon**

-   Ready for:

    -   Log-return modeling
    -   GARCH / ETS
    -   Weekly aggregation
    -   Regime analysis

------------------------------------------------------------------------
